---
title: "RepData_PeerAssessment1"
author: "SaraGordon"
date: "5/24/2021"
output: github_document
---

## Loading and preprocessing the data
Show any code that is needed to:  
- Load the data 
- Process/transform the data (if necessary)

```{r loading, echo = TRUE}
if (!file.exists("data")) {dir.create("data")}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, destfile = "./data/Factivity.zip", method = "curl")
unzip("./data/Factivity.zip", exdir = "./data")

list.files("./data")

ActData <- read.csv("./data/activity.csv", sep = ",", header=TRUE)

##Change the date column to date class
ActData$date <- as.Date(ActData$date)
str(ActData)

```

## What is total number of steps taken per day?
1. Calculate the total number of steps per day

```{r steps1, echo = TRUE}
library(dplyr)
stepSum <- summarize(group_by(ActData, date), stepTotal=sum(steps))
```

2. Make a histogram of the total number of steps per day

```{r steps2, echo = TRUE}
hist(stepSum$stepTotal, breaks = 60, main = "Histogram of Daily Steps", xlab = "Total steps per day")
```

3. Calculate and report the mean/median of the total number of steps per day

```{r steps3, echo = TRUE}
stepMean <- mean(stepSum$stepTotal, na.rm = TRUE)
stepMedian <- median(stepSum$stepTotal, na.rm = TRUE)
```

The mean of the total number of steps is `r stepMean`.  
The median of the total number of steps is `r stepMedian`.

## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r daily1, echo = TRUE}
dailySteps <- summarize(group_by(ActData, interval), avgSteps=mean(steps, na.rm=TRUE))

with(dailySteps, plot(interval, avgSteps, type = "l", ylab = "Average steps", xlab = "Five-minute interval"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r daily2, echo = TRUE}
which.max(dailySteps$avgSteps)
busiest <- dailySteps[104,]
intName <- busiest$interval
intAvg <- busiest$avgSteps
```

The interval with the highest average number of steps is `r intName`, with `r intAvg` steps.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset.

```{r NAs, echo = TRUE}
naSum <- sum(is.na(ActData))
naSum
```

2. Devise a strategy for filling in all of the missing values in the dataset. 

3. Create a new dataset that is equal to the original but with the missing data filled in.

```{r NAs2, echo = TRUE}
library(Hmisc)
NAData <- ActData

#Impute using mode function
ImpData <- NAData %>%
  group_by(interval) %>%
  mutate(steps = impute(steps, median))

#Detach Hmisc so it doesn't interfere with dplyr
detach("package:Hmisc", unload = TRUE)
```


4a. Make a histogram of the total number of steps taken each day.
```{r NAs3, echo = TRUE}
ImpSum <- summarize(group_by(ImpData, date), stepTotal=sum(steps))

hist(ImpSum$stepTotal, breaks = 60, main = "Histogram of Daily Steps, Imputed", xlab = "Total steps per day")
```


4b. Calculate and report the mean and median total number of steps taken per day. 
```{r NAs4, echo = TRUE}
impMean <- mean(ImpSum$stepTotal)
impMedian <- median(ImpSum$stepTotal)
```

The mean of the total number of imputed steps is `r impMean`.  
The median of the total number of imputed steps is `r impMedian`.


4c. Do these values differ from the estimates from the first part of the assignment?

```{r NAs5, echo = TRUE}
if (impMean == stepMean) {
  print("The mean number of steps from the imputed and not-imputed datasets is the same.")
} else if (impMean < stepMean) {
  print("The imputed mean is less than the not-imputed mean.")
} else if (impMean > stepMean) {
  print("The imputed mean is greater than the not-imputed mean.")
}

if (impMedian == stepMedian) {
  print("The median number of steps from the imputed and not-imputed datasets is the same.")
} else if (impMedian < stepMedian) {
  print("The imputed median is less than the not-imputed median.")
} else if (impMedian > stepMedian) {
  print("The imputed median is greater than the not-imputed median.")
}

```

4d. What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r NAs6, echo = TRUE}
ActStepSum <- sum(ActData$steps, na.rm = TRUE)
ImpStepSum <- sum(ImpData$steps)

if (ActStepSum == ImpStepSum) {
  print("The total number of steps from the imputed and not-imputed datasets is the same.")
} else if (ActStepSum < ImpStepSum) {
  print("The imputed step total is less than the not-imputed step total.")
} else if (ActStepSum > ImpStepSum) {
  print("The imputed step total is greater than the not-imputed step total.")
}

```

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” – indicating whether a given date is a weekday or weekend day.

```{r days1, echo=TRUE}
ImpData$dayName <- weekdays(ImpData$date)

weekdata <- subset(ImpData, dayName != c("Saturday", "Sunday"))
weekenddata <- subset(ImpData, dayName == c("Saturday", "Sunday"))

weekdata$dayType <- "weekday"
weekenddata$dayType <- "weekend"

ImpData2 <- rbind(weekdata, weekenddata)

ImpData2$dayType <- as.factor(ImpData2$dayType)
```


2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r days2, echo=TRUE}
library(ggplot2)
weekSteps <- summarize(group_by(ImpData2, interval, dayType), steps=mean(steps))

ggplot(data=weekSteps, aes(x=interval, y=steps)) + facet_grid(dayType~.) + geom_line() + labs(title = "Average Steps, Weekday vs Weekend") + labs(x = "Time of Day") + labs(y = "Average Steps")
```
